Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-10-30 manu-draft.qmd
more results text

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature an precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$. Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.  The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).  The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.  The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay ([@fig-meteowqraw]).  Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).  Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.  As expected, intra-annual variability of water temperature was much higher than salinity following the seasonal variation in temperature ([@fig-meteowqraw]).  This variation may also be related to differences among the monitoring stations in each segment.  Statistically significant increasing temperature trends were also observed for the FIM datasets from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).  No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).  However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.  Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).

@@

An example of mixed effects models for two selected thresholds is shown in @fig-mixeff.

$$
Y_{ij} = \beta_0 + \beta_1X_{ij} + u_i + \epsilon_{ij}
$$

>>>>

Long-term meteorological data from Tampa International Airport showed significantly increasing trends for air temperature and precipitation [@fig-meteowqraw]. Mean annual air temperature has increased by `r airtempslo` $^\circ$C per year (`r airtemppvl`, $R^2$ `r airtemprsq`). Mean annual air temperature in 1975 was `r airtempstr`, whereas current mean annual air temperature in 2022 was `r airtempend`, showing an overall increase in the period of record of `r airtempdif` $^\circ$. Total precipitation has increased by `r prpslo` mm per year (`r prppvl`, $R^2$ `r prprsq`). Total precipitation in 1975 was `r prpstr`, whereas current total precipitation in 2022 was `r prpend`, showing an overall increase in the period of record of `r prpdif` mm. Trends in total annual hydrologic load from 1985 to present followed total annual precipitation, although the linear model was not significant likely due to a shorter period of record.  The SPI showed notable anomalies in precipitation, with pronounced rainy periods in 1980, the late 1990s, 2005, and the last ten years.

Increasing water temperature and decreasing salinity generally followed the meteorological trends for all three *in situ* datasets (EPC, FIM, and PDEM, @fig-meteowqraw, @tbl-temptrndtab, @tbl-saltrndtab, figures in supplement).  The strongest trends were observed for the EPC dataset which had the longest record from 1975 to 2022.  The top and bottom water temperature or salinity changes were similar across bay segments likely because of the relatively shallow water depths in Tampa Bay (@fig-meteowqraw).  Trends in water temperature were similar across bay segments with significant increases varying from 0.3 to 0.4 $^\circ$C per year, equating to 1.4 (OTB) to 2.0 (HB) $^\circ$C total increase from 1975 to present (@tbl-temptrndtab).  Salinity trends were similar between bay segments, although overall salinity was higher for bay segments closer to the Gulf of Mexico. Only Old Tampa Bay and Lower Tampa Bay had significantly decreasing trends, with decreases of 0.06 (OTB) and 0.04 (LTB) ppt per year, totaling an overall decrease of 2.7 and 2.0 ppt from 1975 to present.  As expected, intra-annual variability of water temperature was much higher than salinity following the larger seasonal variation in temperature ([@fig-meteowqraw]).  This variation may also be related to differences among the monitoring stations in each segment.  Significant increasing temperature trends were also observed for the FIM dataset from 1996 to present in OTB and from 2004 to present in HB (@tbl-temptrndtab).  No significant temperature or salinity trends were observed in the other bay segments for the FIM and PDEM datasets from 1996 to present or 2004 present (see figures in supplement).  However, the EPC dataset showed significant increasing temperature across all time periods and bay segments.  Salinity trends from the EPC dataset were not significant from 1996 to present, but were significantly decreasing from 2004 to present in all bay segments (@tbl-saltrndtab).

@@

Linear mixed-effects models showed varying trends in the potential stressor metrics for the number of days when temperature was above, salinity was below, or both occurred relative to different thresholds.  All of the temperature models for each of the three thresholds showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$ (see supplement).  For all bay segments, the estimated slopes for the number of days when temperature was above 30 $^\circ$C were similar among bay segments, varying from 0.9 days per year in LTB to 1.1 days per year in MTB.  Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022.

The salinity models were less similar between bay segments compared to the temperature models (see supplement). None of the bay segments had a significantly increasing number of days when salinity was below 15 ppt.  For MTB, only the number of days when salinity was below 25 ppt was significantly increasing at a rate of 0.67 days per year.  Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB.  The number of days per year when salinity was below 20 ppt in OTB increased by 1 day per year, whereas the number of days in HB increased by 0.5 days per year from 1975 to present.  Likewise, the average number of days when salinity was 20 ppt at the beginning and end of the period of record varied from 11 (HB) to 23 (OTB) days in 1975 to 25 (HB) to 68 (LTB) days in 2022.

Times when both temperature was above a threshold and salinity was below a threshold also varied by bay segment.  None of the models were significant for LTB, four were significant for MTB, seven were significant for HB, and four were significant for OTB.  The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.2, and 0.7 days per year for OTB, HB, and MTB, respectively. Likewise, the average number of days for both varied from the beginning to the end of the period of record from 6, 12, and 1 day(s) in 1975 to 72, 68, and 24 in 2022 for OTB, HB, and MTB respectively.

@fig-mixeff provides visual examples of the mixed-effects models for the estimated number of days over time for each bay segment when temperature was above 30$^\circ$, salinity was below 25 ppt, and when both occurred. These thresholds were chosen moving forward because they are slightly more conservative than those with the largest slopes (i.e., higher temperature threshold), while still showing significant increases over time (excluding LTB).  Temperature trends were similar among segments, whereas the number of days when salinity was below the threshold varied by proximity to the Gulf of Mexico (Table 3).  The number of days when both temperature was above and salinity was below the threshold generally followed the trends for the number of days when salinity was below the threshold, although the slopes varied.  These thresholds were used for comparison to seagrass changes described below, based primarily on the statistical strength of the trends and the variance of counts among bay segments.  That is, more restrictive thresholds did not provide sufficient counts of days per year to develop models and the chosen thresholds were based primarily on statistical considerations rather than ecological considerations, as discussed below.
